A Method for the spectroscopic inference of fundamental stellar parameters

### 2014-04-30 paper/paper.tex
Caption error
Revisions and added text to paper

\documentclass[preprint]{aastex} %double-column, single-spaced document:
%\documentclass[iop,floatfix]{emulateapj}

\usepackage{hyperref}
%\usepackage{graphicx}
%\usepackage{apjfonts}
\usepackage{enumerate}
\usepackage{amsmath,amssymb}
\usepackage{geometry}
\usepackage{bm}

\newcommand{\prob}{{\rm prob}}
\newcommand{\qN}{\{q_i\}_{i=1}^N}
\newcommand{\qM}{\{q_{im}\}_{i=1,m=0}^{N,M}}
\newcommand{\yN}{\{y_i\}_{i=1}^N}
\newcommand{\vt}{\vec{\theta}}
\newcommand{\vg}{\vt_{\star, {\rm grid}}}
\newcommand{\vpp}{\vt_{\star, {\rm post}}}
\newcommand{\vstar}{\vt_{\star}}
\newcommand{\vN}{\vt_{\rm N}}
\newcommand{\vc}{\vec{c}}
\newcommand{\fM}{ {\bm M}}
\newcommand{\fMi}{M_i}
\newcommand{\fD}{ {\bm D}}
\newcommand{\fDi}{D_i}
\newcommand{\dd}{\,{\rm d}}
\newcommand{\trans}{\mathsf{T}}
\newcommand{\Z}{[{\rm Fe}/{\rm H}]}
\newcommand{\A}{[\alpha/{\rm Fe}]}

@@

\title{Methods for the spectroscopic inference of fundamental stellar parameters}
\author{\today{}\\
\medskip
Ian~Czekala\altaffilmark{1} et al.
%Author2\altaffilmark{2},
}

@@

\section{Introduction}
The fundamental properties of stars---effective photospheric temperature, surface gravity, and metallicity--are crucial for our understanding of pre-main sequence stellar evolution \citep{dm97, bca+02}. The exoplanet community is also intensely focused on deriving accurate stellar properties because the properties of newly discovered exoplanets must necessarily be quoted relative to those of their host stars (e.g. \citealt{tfs+12,blj+12,ssm+13}) and uncertain star properties translate into uncertain planet properties \citep{kan14}. Many techniques\footnote{If you are interested, I have also compiled a more verbose review of modern techniques described in the literature. However, for the sake of brevity I have omitted it from this document.} exist for stellar characterization such as line indexes, machine learning, and spectral synthesis\footnote{Given infinite computational power, we would prefer to fit \emph{all} of the parameters that completely describe a synthetic spectrum: atmospheric structure, individual elemental abundances, and atomic constants describing the opacity contribution of a line. Approximations to the synthesis and efficient MCMC sampling parallelized on a cluster might soon put this approach within reach.}. In order to derive accurate and robust stellar properties for our small sample ($\sim 20$) of T Tauri stars (some binary), we have decided to forward model the observed spectra by drawing from a grid of synthetic spectra.

\begin{table}[!htb]
\begin{tabular}{ll}
\hline
\hline
Symbol & Description\\
\hline
\hline
$i$ & index specifying a pixel\\
$\lambda_i$ & wavelength corresponding to a given pixel $i$\\
$\vg$ & fundamental stellar parameters, $T_{\rm eff}, \log(g), \Z, \A$\\
  & that parameterize a synthetic spectrum from the grid\\
$\vpp$ & stellar parameters $v \sin i$, $v_z$, $A_V$, and $R^2/d^2$ that\\
  & are applied during ``post processing'' of the synthetic spectrum\\
$\vstar$ & $\{\vg,\vpp \}$\\
$c_0, c_1, \ldots$ & Chebyshev polynomial coefficients for residual flux calibration\\
$\vN$ & the set of nuisance parameters composed of $\{c_0, c_1, \ldots, c_N\}$\\
$k(i | \vN)$  & polynomial function that multiplies a synthetic spectrum\\
$\vt$ & the parameters $\{\vg, \vpp, \vN\}$ that completely describe a model spectrum\\
$\fDi$ & data flux for a given pixel, $D(\lambda_i)$\\
$\fD$ & data vector comprised of all $\fDi$, $i = \{0, \ldots, N\}$\\
$\fMi$ & model flux for a given pixel, $M(\lambda_i | \vt)$\\
$\fM$ & model vector comprised of all $\fMi$, $i = \{0, \ldots, N\}$\\
$\sigma_i$ & Poisson noise for a given pixel $i$\\
$R_i$ & residuals $(\fDi - \fMi)/\sigma_i$\\
\hline
\end{tabular}
\caption{Nomenclature used in this document}
\label{tab:nomenclature}
\end{table}

\subsection{Pixel to pixel techniques using synthetic spectral libraries}
\label{sec:pix}
Most forward modeling techniques compare pixels directly using some form of
\begin{equation}
  \chi^2 = \sum_i \left [\frac{ \fDi - \fMi}{\sigma_i} \right ]^2 = \sum_i R_i^2
\end{equation}
where the variables are defined in Table~\ref{tab:nomenclature}. A serious problem with this approach is that synthetic spectra are not perfect: systematic errors invalidate the $\chi^2$ assumption and bias parameter estimates and their uncertainties.

Current methods either circumvent this problem by using high quality synthetic grids that were painstakingly tuned over a narrow wavelength range \citep[$\lesssim 500$ \AA]{blj+12,sb13}, or they use less-perfect grids with full optical coverage and sigma clip the bad regions of the spectrum \citep{kpb+09,mga13}. Masking (either implicitly or explicitly, respectively) is troublesome because, in the words of \citet{mga13}, ``regions with [only] modest agreement between the real and synthetic spectra [that might be masked] may contain more temperature information than regions with slightly better matches, and we have no a priori information about what spectral regions are the most temperature sensitive.'' Masking could very well destroy stellar information in the pursuit of a better fit.

\begin{figure}[!htb]
\begin{center}
\includegraphics{lineclasses}
caption{The four classes of line behavior. \textbf{Class 0}: the model behaves as expected. \textbf{Class I}: a spectral line is present in the data but missing from the model (in this case, the line is also blended). \textbf{Class II}: a spectral line exists in the model where there is only continuum in the data. \textbf{Class III}: lines exist in both the data and model, but some are too strong while others are too weak. The lower panels show the residual spectra in units of the Poisson uncertainty. The data is a spectrum of WASP-14, an F star. The model is from the PHOENIX grid\citep{hwd+13}.}
\label{fig:lineclasses}
\end{center}
\end{figure}

\subsection{Classes of synthetic spectral line behavior}
Illustrated in Figure~\ref{fig:lineclasses} are three ways that synthetic spectra often diverge from reality. When a spectral line appears in the data but not in the model, due to interstellar or telluric absorption or a missing opacity source, we call this a \textbf{Class I} error. When a spectral line appears in the model but not in the data, presumably due to a combination of incorrect atomic constants or atmospheric structure, we label it a \textbf{Class II} error. Both of these types of errors are present over a wide range of probable stellar parameters (1000 K in $T_{\rm eff}$, and 1 dex in $\log(g)$, $\Z$, and $\A$) and there does not exist a $\vg$ that could ever fit these lines, making these regions useless for determining stellar parameters. We know that for a normal AFGK star Class I residuals will always be negative and will not change significantly with $\vg$, while Class II residuals will always be positive and their magnitude should decrease with decreasing metallicity. To help identify Class I and II errors,  we could use the partial derivatives of an individual synthetic spectrum with respect to $\vg$ and a prior on the range of parameters over which the residuals will be correlated to determine which lines show Class I or II behavior.

When some line strengths do not match at a $\vg$ that properly fits most other lines, we term these \textbf{Class III} errors. These types of errors are pernicious because there does exist a $\vg$ that would fit each line individually, but no one $\vg$ can be chosen that fits all the lines. We must then decide which regions to trust and by how much, because Class III regions \emph{still contain information about $\vg$}. In fact, they might contain \emph{the most} information about $\vg$ since these delicate regions are the most sensitive to changes in $\vg$ and are consequently whey they are the hardest to model correctly. In other words, Class III regions might be where the spectrum gradient ($\sum_{j} \partial \fMi/ \partial \vt_{ \star {\rm grid}, j}$) is the largest.

\section{In the limit of a perfect stellar model}
To generate a model spectrum $\fM(\vg, \vpp)$, we use $\vg$ to linearly interpolate a synthetic spectrum from the grid, and then use $\vpp$ to apply instrumental and rotational broadening in the Fourier domain, Doppler shift, correct for extinction, and then downsample to the exact pixels of the data spectrum using spline interpolation\footnote{Because we are dealing with flux-calibrated data and synthetic spectra, we \emph{resample} to lower resolution. We do not \emph{rebin} because we are dealing with flux-density, not photon counts}.

To account for imperfect flux calibration, we extend the approach of \citet{elh+06} and multiply $\fM(\vg, \vpp)$ by $k(\vN)$, which is a Chebyshev polynomial described by a set of $N$ Chebyshev coefficients. We can either sample in the coefficients or analytically marginalize over them. We favor this approach over the traditional technique of normalizing to the stellar continuum due to concerns about correctly placing the pseudo-continuum in M stars. From now on we refer to $\fM(\vt) = k(\vN) \fM(\vg, \vpp)$. The mathematical details are in \S~\ref{sec:chebyshev}.

Currently we are sampling this $\chi^2$ posterior with the MCMC ensemble sampler \citep{gw10} implemented in \texttt{Python} by \citet{fhl+12} as \texttt{emcee}. In this section, we described our approach to sample the posterior assuming perfect synthetic spectra. However, if one does a straight up $\chi^2$ fit to our spectra, one can get posteriors that have 1-sigma contours of $\pm 1$ K and $\pm 0.01$ dex in gravity and metallicity. This would be as precise as stellar parameters obtained with astroseismology! Clearly there is something going wrong here.

\section{Beyond a perfect stellar model}
\label{sec:residuals}
\subsection{Additional sources of noise}
\begin{itemize}
  \item CCD readnoise and imperfect 1D spectrum extraction over night sky lines will almost definitely increase the noise in a pixel beyond theoretical Poisson limits. Normally this information will be encapsulated in a ``sigma spectrum'' output by IRAF, but unfortunately we do not have access to this for our current dataset.
  \item The stellar continuum can flicker/jitter (as shown by Kepler), which would produce a small departure from a perfect continuum. This might explain why even in the Class 0 region of our spectrum (Figure~\ref{fig:lineclasses}, left panel) the residuals are slightly larger than 1 $\sigma$ on average. For our T Tauri stars this effect may be more pronounced.
  \item When we interpolate a synthetic spectrum from the library to a specific $\vg$, we will introduce interpolation errors depending on how close we are to a grid point in the library. \citet{hus12} investigate the possible range of interpolation errors at specific $\vg$ by dropping out a spectrum from the grid, interpolating over it, and then comparing the true spectrum to the interpolated spectra. For hotter stars with well defined continua, we find the maximum error is $\lesssim 5$\%, while for M stars the error could be as large as $\sim 20$\% for some spectral features. By doing this interpolation test across the grid, we could determine an ``interpolation error spectrum'' that would encapsulate the maximum error possible for an interpolated spectrum, weighted by its distance from the nearest grid points. If a typical spectrum has signal/noise of $\sim 30$ per pixel (or noise/signal ratio of $1/30 \approx 3$\%), an interpolation noise contribution of $\sim 5$\% the level of the signal level would be significant. 
\end{itemize}
However, we have tried arbitrarily increasing the Poisson errors by a factor of 2 or 3 and it didn't significantly change the results, suggesting that something more drastic is needed beyond simply adding in more noise.

\subsection{Gallery of likelihood functions}
\label{sec:gal}
A problem with the $\chi^2$ likelihood function is that the Gaussian tails rapidly decline, making outliers extremely improbable. A common approach for ``robust regression'' is to use a likelihood function with fatter, more forgiving, tails. Although not rigorously motivated, this is can be a quick thing to try. We explored several likelihood functions including a combination of Gaussian with exponential wings, a Lorentzian (Cauchy) function, and a function described  by \citet{ss06}, which treats the noise estimate for each data point suspiciously. In Sivia's model, the true noise $\sigma$ can be no smaller than some original estimate $\sigma_0$, but it may be much larger. By putting a prior on $\sigma$, 
\begin{equation}
  p(\sigma_i | {\sigma_0}_i) = \left \{ \begin{array}{cc}
    {\sigma_0}_i/\sigma_i^2 & \sigma_i  \leq {\sigma_0}_i\\
    0                 & \textrm{else}
  \end{array}
    \right.
\end{equation}
and marginalizing over the $\chi^2$ likelihood, the new likelihood function becomes
\begin{equation}
p(\fDi | \fMi, {\sigma_0}_i) = \frac{1}{ {\sigma_0}_i \sqrt{2 \pi}} \left [ \frac{1 - e^{-R_i^2/2}}{R_i^2} \right ] 
\end{equation}
where $R_i = (\fDi - \fMi)/{\sigma_0}_i$. This function has even fatter tails than the exponential model and seems to give results less biased by outliers, however it is not truly satisfying. Which lines contributed most to the posterior? What about the wings of class I and II errors? These individual pixels might still be given a weight of ${\sigma_0}_i$ because they are not nearly as bad as the line centers.

Other functions we might try in the future include a Student-$t$ distribution, a linear combination of Gaussians, or the $\chi^2_Q$ function described in \citet{hbl10}. While ``softer'' likelihood functions do help somewhat, we would like to move beyond tweaking the fitting function towards an approach that properly incorporates our knowledge of the systematic errors of the synthetic grids. In the remaining two sections of this document we outline potential ideas that we might try next, but have not yet implemented in code.

\section{Bad data mixture model}
The bad data model \citep{pre97,hbl10} is a powerful framework for forward modeling data, which allows the modeler to explicitly account for outliers using a generative model. In the simplest mixture model, each pixel has a binary flag $q_i$ which specifies whether that data point is good or bad. Then, for the entire data set we have an ensemble of $\qN$. If this ensemble of $\qN$ is chosen appropriately, and the good pixels are flagged as good and the bad pixels are flagged as bad, the posterior will be maximized. We adopt $P_B$ for the \emph{a priori} probability that a given pixel will be bad, $Y_b$ for the mean flux of the bad pixels, and $V_b$ for their variance. Written out explicitly, this is
\begin{multline}
  {\cal L} = p({\bm D} | \vt,\qN, Y_b,V_b) = \prod_{i=1}^N \left [ \frac{1}{\sqrt{2 \pi} \sigma_i} \exp \left[ - \frac{(\fDi - \fMi)^2}{2 \sigma_i^2} \right] \right ]^{q_i}  \times \\
  \left [ \frac{1}{\sqrt{2 \pi (V_b + \sigma_i^2)}} \exp \left[ - \frac{(\fDi - Y_b)^2}{2 (V_b + \sigma_i^2)} \right]
  \right ]^{1 - q_i}
\end{multline}
If we put a Bernoulli prior on the $q_i$
\begin{equation}
  p(q_i | P_b) = [1-P_b]^{q_i} P_b^{[1-q_i]}
\end{equation}
the probability of any given ensemble of $\qN$ is 
\begin{equation}
  p(\qN | P_b) = \prod_{i=1}^N [1-P_b]^{q_i} P_b^{[1-q_i]}
\end{equation}
We can analytically marginalize over the settings for a single flag by a discrete sum over the two states of $q_i = 1$ and $q_i = 0$. 
\begin{multline}
  p(\vt, P_b, Y_b, V_b | D_i)  =  \int p( \vt, q_i, P_b, Y_b, V_b| D_i) {\rm d} q_i \\
  = p(\vt, q_i=1, P_b, Y_b, V_b | D_i) + p(\vt, q_i=0, P_b, Y_b, V_b | D_i) \\
  = \Bigl[ p_{\rm good}(D_i | \vt)\, p(q_i=1 | P_b) +  p_{\rm bad}(D_i | Y_b, V_b)\, p(q_i=0 | P_b)\Bigr] \times p(\vt ,P_b, Y_b,V_b) 
\end{multline}

\begin{multline}
  p(\vt,P_b,Y_b,V_b|\,{\bm D}) \propto \prod_{i=1}^N \Bigl [ [1 - P_b] p_{\rm good}(D_i |\vt)  + P_b p_{\rm bad}(D_i | Y_b, V_b)\Bigr ] \times p(\vt,P_b, Y_b, V_b)\\
  \boxed{
    = \prod_{i=1}^N \left\{ \frac{1 - P_b}{\sqrt{2 \pi \sigma_i^2}} \exp \left[ - \frac{(\fDi - \fMi)^2}{2 \sigma_i^2} \right] + \frac{P_b}{\sqrt{2 \pi (V_b + \sigma_i^2)}} \exp \left[- \frac{(\fDi - Y_b)^2}{2 (V_b + \sigma_i^2)} \right] \right\}
  \times p(\vt,P_b,Y_b,V_b)}
\end{multline}

In its simplest form, the bad data model results in a likelihood function that is a linear combination of Gaussians. Although this functional form might be similar to some of the likelihood functions in \S~\ref{sec:gal}, the benefit of this bad data approach over a gallery of likelihood functions is that it is motivated by the characteristics of the bad data ($P_B$, $Y_b$, and $V_b$). If we wish to add additional information about the characteristics of the bad data, we can do so.

\subsection{Mixture of multiple components}
When we believe that a sample population might be composed of several populations with different behaviours (Class 0, I, II, and III), we can expand the bad data model to a multiple mixture model. Following \citet{gcs+04}, we develop a mixture model with $M=4$ components, one for each class of error. Our likelihood function is then
\begin{equation}
  p(\fD, \vec{Q} | \vt, J) = \prod_i^N \prod_{m=0}^M \left [ J_m {\cal L}_m(\fDi | \vt_m) \right ]^{q_{im}}
\end{equation}
where $\vec{Q} = \qM$ is a two dimensional array ($N \times M$) of all the individual $q_{im}$ flags, and $J_m$ is a weight that represents the total fraction of each component in the mixture\footnote{The weights are normalized $\sum_m J_m = 1$}, and ${\cal L}_m$ is the likelihood function for $m$th mixture component. At this point, $\vt$ now contains all of the parameters that might be used by any of the likelihood functions. $\vt_m$ is used to denote only those parameters that will be used for ${\cal L}_m$. Given what we know about these synthetic spectral errors, we could specify the following likelihood functions
\begin{equation}
  \begin{array}{lcccc}
 {\cal L}_0 &= p(\fDi, q_{i0} = 1) &\rightarrow & \textrm{Class 0} & \chi^2\\
 {\cal L}_1 &= p(\fDi, q_{i1} = 1) &\rightarrow & \textrm{Class 1} & \textrm{flat lnprob}\\
 {\cal L}_2 &= p(\fDi, q_{i2} = 1) &\rightarrow & \textrm{Class 2} & \textrm{flat lnprob}\\
 {\cal L}_3 &= p(\fDi, q_{i3} = 1) &\rightarrow & \textrm{Class 3} & \textrm{soft } \chi^2\\
\end{array}
\end{equation}
This would treat Class I and II errors as if they had been masked from the dataset but allow Class III to have a slight influence on the posterior since these regions still contain information about $\vstar$. The softening term of the $\chi^2$ could even be sampled for, much like the $V_b$ of the original bad data model.

\subsection{Spatial correlations of pixels}
Previously we assumed that each data point is independent, but we know that the spectrum is spatially correlated across several pixels with a correlation length set by the width of the convolution of the stellar rotation kernel ($v \sin i$) and the instrumental resolution kernel, which is also the typical width of a spectral line\footnote{Some very strong lines (such as the Balmer lines or Mg b triplet) will have wings with a larger correlation length.}. This correlation length will have some dependence on the spectral type of star. For example, F and G stars will likely have singular, well-defined lines while M stars might have a large molecular feature covering a wide swath of the spectrum. Neighboring pixels are actually correlated in two coordinates: the $q_{im}$ flags and the residuals $R_i$.

\subsubsection{Correlation of the $q_i$ flags}
The pixels at the center of a bad line will give a strong residual response, which will likely be flagged as bad. However, the pixels in the wings of the spectral line might not be sufficiently egregious to be properly flagged. To remedy this situation, we attempt to develop a bad data model which includes spatial (wavelength) correlations.

>>>>

%\documentclass[preprint]{aastex} %double-column, single-spaced document:
\documentclass[iop,floatfix]{emulateapj}

\usepackage{hyperref}
%\usepackage{graphicx}
%\usepackage{apjfonts}
\usepackage{enumerate}
\usepackage{amsmath,amssymb}
\usepackage{geometry}
\usepackage{bm}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\newcommand{\prob}{{\rm prob}}
\newcommand{\qN}{\{q_i\}_{i=1}^N}
\newcommand{\qM}{\{q_{im}\}_{i=1,m=0}^{N,M}}
\newcommand{\yN}{\{y_i\}_{i=1}^N}
\newcommand{\vt}{\vec{\theta}}
\newcommand{\vg}{\vt_{\star, {\rm grid}}}
\newcommand{\vpp}{\vt_{\star, {\rm post}}}
\newcommand{\vstar}{\vt_{\star}}
\newcommand{\finst}{f_{\lambda, {\rm inst}}}
\newcommand{\fsynth}{f_{\lambda, {\rm synth}}}
\newcommand{\vN}{\vt_{\rm N}}
\newcommand{\vc}{\vec{c}}
\newcommand{\fM}{ {\bm M}}
\newcommand{\fMi}{M_i}
\newcommand{\fD}{ {\bm D}}
\newcommand{\fDi}{D_i}
\newcommand{\fR}{ {\bm R}}
\newcommand{\dd}{\,{\rm d}}
\newcommand{\trans}{\mathsf{T}}
\newcommand{\Z}{[{\rm Fe}/{\rm H}]}
\newcommand{\A}{[\alpha/{\rm Fe}]}

\newcommand{\todo}[1]{ \textcolor{Blue}{\\TODO: #1}}

%% Nomenclature guide
%%%%%%%%%%%%%%%%%%%%%
% * collections of synthetic spectra are generically called ``libraries.'' The term 
% ``grid'' is used when referring specifically to the spacing of spectral parameters
% in the library

@@

\title{A Method for the spectroscopic inference of fundamental stellar parameters}
\author{\today{}\\
\medskip
Ian~Czekala\altaffilmark{1} et al.
%Author2\altaffilmark{2},
}

@@

\section{Introduction}
\begin{itemize}
  \item Where our technique fits in the ecosystem of stellar methods.
  \item Applicable to any kind of spectrum (long-slit, echelle, infrared, flux-calibrated or not)
  \item Examples of fields where accurate and unbiased stellar parameters are crucial. Exoplanets. T Tauri stars.
\end{itemize}

\section{Method}
Typically, the stellar properties we most desire are effective temperature $T_{\rm eff}$, radius or surface gravity $\log g$, and metallicity $\Z$. Several high-quality libraries of synthetic spectra now exist (PHOENIX, Kurucz, more being developed for GAIA) and are parameterized by these fundamental stellar parameters, which together, we call $\vg$. Synthetic libraries provide high-resolution spectra which span a range of $\vg$ covering the main sequence of spectral types. The library is typically specified on a grid of equal spacing in $T_{\rm eff}$, $\log g$, and $\Z$.

In addition to these fundamental parameters, a star also has several observed properties that are a function of its kinematics, geometry, and location in our galaxy: projected rotational velocity $v \sin i$, line of sight velocity $v_z$, total extinction $A_V$, and solid angle $\Omega$. We call these secondary parameters $\vpp$, because we can model their effects on the synthetic spectra in a ``post-processing'' step by convolution with an appropriate kernel or multiplication with a smooth function. Together, we call $\vstar = \{\vg, \vpp\}$.

Using a synthetic stellar library as a backend and applying post-processing techniques, we attempt to forward-model the observed stellar spectra in order to determine the best-fit $\vstar$. Because properly sampled spectra have correlated pixels, we present a framework which self-consistently models these correlations.

\begin{table}[!htb]
\begin{tabular}{ll}
\hline
\hline
Symbol & Description\\
\hline
\hline
$i$ & index specifying a pixel\\
$\lambda_i$ & wavelength corresponding to a given pixel $i$\\
$\vg$ & fundamental stellar parameters, $T_{\rm eff}, \log(g), \Z, \A$\\
  & that parameterize a synthetic spectrum from the grid\\
$\vpp$ & stellar parameters $v \sin i$, $v_z$, $A_V$, and $R^2/d^2$ that\\
  & are applied during ``post processing'' of the synthetic spectrum\\
$\vstar$ & $\{\vg,\vpp \}$\\
$\finst(\lambda)$ & data spectrum\\
$\fsynth(\lambda)$ & synthetic spectrum\\
$c_0, c_1, \ldots$ & Chebyshev polynomial coefficients for residual flux calibration\\
$\vN$ & the set of nuisance parameters composed of $\{c_0, c_1, \ldots, c_N\}$\\
$k(i | \vN)$  & polynomial function that multiplies a synthetic spectrum\\
$\vt$ & the parameters $\{\vg, \vpp, \vN\}$ that completely describe a model spectrum\\
$\fDi$ & data flux for a given pixel, $D(\lambda_i)$\\
$\fD$ & data vector comprised of all $\fDi$, $i = \{0, \ldots, N\}$\\
$\fMi$ & model flux for a given pixel, $M(\lambda_i | \vt)$\\
$\fM$ & model vector comprised of all $\fMi$, $i = \{0, \ldots, N\}$\\
$\sigma_i$ & Poisson noise for a given pixel $i$\\
$R_i$ & residuals $\fDi - \fMi$\\
$\fR$ & residual vector $\fD - \fM$\\
\hline
\end{tabular}
\caption{Nomenclature used in this document}
\label{tab:nomenclature}
\end{table}

\subsection{Interpolation from synthetic spectral libraries}
Creating a new spectrum with a specific $\vg$ requires either synthesizing a new spectrum using radiative transfer codes or interpolating from nearby grid points in the library. Because it is computationally expensive to synthesize a spectrum at high resolution over a wide wavelength range, we choose to interpolate. Spline interpolation for spectra \citep{hus12}.
\todo{Implement band-limited interpolation from the synthetic grid using spline interpolation. Caution: As is, 100K/0.5 logg spacing in $\vg$ may not be Nyquist sampling.}
\todo{Investigate Tom Loredo's mentions of ``Cosmological Bayesian Emulator'' for using Gaussian processes for interpolating from a simulation grid.}

\subsection{Post-processing synthetic spectra to match the data}
Libraries of synthetic spectra are generally computed at high resolution ($R \geq $100,000), sampled with many pixels per resolution element, and do not include any rotational broadening or account for any instrumental effects. In order to transform a raw synthetic spectrum interpolated from the grid into one that matches the spectrum of a real star, we must post-process the spectrum to account for these secondary effects. The projected rotational velocity of the star, parameterized by $v \sin i$, broadens spectral line profiles and is mathematically described by convolution with a kernel $g_{v \sin i}(\lambda)$. UV, optical, and infrared spectra are acquired using a spectrograph which disperses light onto a CCD, defining a specific resolution and sampling rate for the spectrum. Because the resolution of the spectrograph and spacing of the pixels are different from the synthetic spectral library, we must convolve the raw spectrum with the line spread function (LSF) of the instrument $g_{\rm LSF}(\lambda)$ and resample to the exact pixels of the CCD. In wavelength space, the $v \sin i$ and LSF operations are represented by the convolution of the synthetic spectrum with these two kernels
\begin{equation}
  \finst(\lambda) = g_{v \sin i}(\lambda) \otimes g_{\rm LSF}(\lambda) \otimes \fsynth(\lambda)
  \label{eqn:convolution}
\end{equation}
Using the convolution theorem, we can rewrite these operations as multiplications in Fourier space 
\begin{equation}
  F_{\lambda, {\rm inst}}(s) = G_{\rm LSF}(s) G_{v \sin i}(s) F_{\lambda, {\rm synth}}(s)
\end{equation}
where $G \leftrightarrow g$ and $F \leftrightarrow f$ are Fourier transform pairs. In order to use the fast Fourier transform (FFT) to execute these operations using discrete samples of the synthetic spectrum $\fsynth$, we must first resample $\fsynth$ so that it is on a uniform grid. In the case of a spectrum, the natural uniform grid is one that is equally sampled in velocity space, such there is an equal velocity shift $\Delta v$ between pixels. This results in a wavelength grid that is linearly sampled in the logarithm of wavelength. Therefore, the Fourier coordinate $s$ is the number of cycles per sampling interval, having units of inverse velocity $[{\rm s}/{\rm km}]$. Next, we multiply the Fourier transform of the synthetic spectrum by the Fourier transforms of the rotational velocity and line spread function kernels. Lastly, we do an inverse FFT to transform the modified spectrum $F_{\lambda, {\rm inst}}$ back to wavelength space, where it is sampled at the wavelength locations of the pixels in the detector ($\finst$). When resampling the spectrum, we are careful to ensure band-limited interpolation using splines in order to prevent introducing non-physical high frequency structure into the spectrum.

Finally we apply corrections for interstellar extinction and Doppler shift. Interstellar extinction attenuates the spectrum by an amount $A_\lambda$, which is wavelength-dependent. We use the (name here) dust-extinction model, which is parameterized by extinction in the $V$ photometric band, $A_V$. We correct for a radial velocity difference $v_z$ between the star and the earth by Doppler shifting the model spectrum.

\todo{Convolution in wavelength space is a possibility. Allows LSF to change with wavelength.}
Pixel width effects are not important as long as the LSF of the instrument is adequately sampled ($\gtrsim 3$ pixels across the LSF). If our spectrum were not properly sampled, we would need to additionally convolve with a boxcar the width of the pixels.

\todo{Doppler shifting the wavelengths could be implemented as a phase-shift in Fourier space, but it's pretty fast as-is.}

Due to small imperfections in the flux-calibration of the data spectrum, there may be broad regions of the spectrum which are offset from the true continuum level by a slight amount. A traditional approach to address this problem is to normalize both the data and the spectrum to the continuum level before comparison. While this may work well for hotter stars with well-defined continuua, for cooler stars with large molecular features, normalization may introduce artificial features into the spectrum when the pseudo-continuum is incorrectly placed. Instead, we choose to multiply the model spectrum by a low-order Chebyshev polynomial that is normalized to unity. The coefficients of the polynomial, $c_0, c_1, \ldots$, will be free-parameters in our model that we will solve for. If the behavior of the spectrograph is characterized by observations of spectrophotometric standards, then we can put priors on the coefficients that will prevent overfitting of the polynomials.

With more uncertainty, the polynomial multiplication can also be used as a substitute for flux-calibration since the sensitivity function of an instrument is usually a smooth function with wavelength. When spectra are not flux calibrated, there is greater uncertainty about the properties of the spectrograph and wider priors must be used on the Chebyshev coefficients. In this case, the polynomial might destroy broad-scale information present in the spectrum by fitting it out, something that would be avoided with more accurate priors provided by the flux calibration for the model.

\subsection{Likelihood function}
To evaluate which parameters of the model $\fM$ fit the data set $\fD$ best we use a standard multidimensional Gaussian likelihood function which allows for covariance between data points. If the residuals vector is 
\begin{equation}
  \fR(\vstar) = \fD - \fM(\vstar)
\end{equation}
with $N$ data points, then the likelihood function is
\begin{equation}
  p(\fD | \vstar) = \frac{1}{\sqrt{(2 \pi)^N \det(C)}} \exp\left ( -\frac{1}{2} \fR^T C^{-1} \fR \right ) 
\end{equation}
and the log-likelihood function is
\begin{equation}
    \ln[p(\vstar)] = -\frac{1}{2} \fR^T C^{-1} \fR - \frac{1}{2} \ln \det C  - \frac{N}{2} \ln 2 \pi 
    \label{eqn:lnprob}
\end{equation}
$C$ is the covariance matrix which describes the covariance between pixels in the spectrum. For independent noise with noise per pixel $\sigma_i$, this matrix is
\begin{equation}
  C = 
  \begin{bmatrix}
    \sigma_0^2 & 0  & \cdots & 0\\
    0 & \sigma_1^2 & \cdots & 0\\
    \vdots  &   & \ddots  & \vdots \\
    0 & 0 & \cdots & \sigma_N^2\\
  \end{bmatrix}
\end{equation}
and Equation~\ref{eqn:lnprob} reduces to the familiar $\chi^2$ form of a sum over the square of the residuals, weighted by the inverse variance of each data point
\begin{equation}
  \ln[p(\vstar)] \propto - \frac{1}{2} \chi^2 = - \frac{1}{2} \sum_i \frac{R_i^2}{\sigma_i^2}
\end{equation}
However, residuals from a spectroscopic fit are often correlated due to the convolution of the stellar rotation and LSF kernels. Additionally, systematic errors in the synthetic spectra (such as incorrect line strengths) will resulted in regions of highly correlated residuals, which requires that we use a covariance matrix with off-diagonal terms $\sigma_{ij}$ and Equation~\ref{eqn:lnprob}, which includes a matrix product instead of a sum over independent pixels.

In order properly account for these correlated residuals, we must understand the covariance structure and parameterize it in our model. We seek to understand two types of covariance structure: the large scale global covariance structure that results from the stellar rotation and LSF kernels, and a regional covariance structure that is localized to areas where spectral line mismatch results in small patches of highly correlated residuals.

\todo{Tom Loredo was saying that if we have a model that perfectly fits the data then our residuals will be uncorrelated, regardless of the PSF/LSF. But is this true? Shouldn't all noise be correlated on some scale by the LSF? Correspond by email to clarify and show an example.}

To account for the global covariance structure, we use a Matern 3/2 kernel.
\todo{Explore difference between using 3/2 Matern kernel and squared exponential kernel.}
\begin{equation}
  k_{\nu = 3/2}(r, a, l) = a \left(1 + \frac{\sqrt{3} r}{l} \right ) \exp \left (- \frac{\sqrt{3} r}{l} \right )
\end{equation}

tapered for compact support using a Hann window with width $r_0 = 6 l$ (Why? Because this looked good).

Specific line covariance. Non-stationary covariance kernel.
\begin{equation}
  k_{\rm G}(x, x^\prime | a, \mu, \sigma) = \frac{a^2}{2 \pi \sigma} \exp \left ( - \frac{[(x - \mu)^2 + (x^\prime - \mu)^2]}{2 \sigma^2}\right )
\end{equation}
\begin{equation}
k(x, x^\prime | h, a, \mu, \sigma) = \exp \left ( \frac{-( x - x^\prime)^2 }{2 h^2} \right ) k_{\rm G}(x, x^\prime | a, \mu, \sigma)
\end{equation}
$h$ is a "bandwidth." If $h$ is small, then there will be high-frequency structure. If $h$ is large, then only low-frequency structure will remain.

These are all parameters that we can sample in. Also, there would be a set of these parameters for each line in the model.

The benefit mainly comes from simply modelling these systematic residuals to begin with, since a model with covariance is far more likely than forcing the fit, and far more justified than masking regions which do not fit.

\subsection{Applications}
Learnt covariance structure can be used to correct the models.
